Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-18 response-to-reviews.Rmd

l258 under soil wet conditions?

*doing this because I don't think we want to talk about soil conditions. I've always had atmospheric conditions in my mind as I've thought about this.*

L298-299 Far red to red light: I think this could be developed here or somewhere else. You covered the vertical variation in light quantity inside the canopy, but you could also report the vertical change in light quality (light spectrum) as the leaves absorb preferentially some wavelengths but reflects or transmit others.

**Added the following lines:**
**Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers. Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000]**

@@

- **Conclusions section, first par (text specifically cited in the comment above): "Similarly, much remains to be learned about how crown exposure influences the temperature sensitivity of woody stem growth. While the preponderance of available data suggest that large canopy trees are the most vulnerable to warming when water is limited, far less is known about the responses of understory trees, which might be more vulnerable to chronic warming stress under relatively mesic conditions (Fig. 4)."**

### Details

There are a few dense sentences or run-ons that could be revised. Lines 192-196 and lines 201-204, for instance. Please revise.

**We revised the following lines:**

- **Lines 192-196: Notably, this pattern can be reversed in open forests, where below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$ due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers, analogous to a ‘canopy greenhouse effect’**

- **Notably, this pattern can be reversed in open forests. Analogous to a 'canopy greenhouse effect', in open forests, below-canopy maximum $T_{air}$ can be warmer than canopy $T_{air}$. This is due to turbulent air mixing and the thermal radiative flux from the soil and the canopy that is intercepted by lower-canopy layers**

- **Lines 201-204: $T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes (Drake et al., 2020; Perez & Feeley, 2020; Michaletz et al., 2015)**

- **$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (section 1) and by leaf traits and stomatal conductance (see sections 3-4). Leaves adjust to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes **

@@

line52: 'with decreases being more commonly documented across the world’s forests'

line128: '… but understory leaf area density is often relatively high in the understory as well' sounds redundant. Please reword.

**We removed the second 'understory' word here:** 
**Tropical and temperate forests with dense canopies dominated by broadleaf trees generally have highest leaf area density in the upper canopy layers, but understory leaf area density is often relatively high ~~in the understory~~ as well**

@@

line140 'Light … decreases from the canopy top to the forest floor' – why isn’t foliage clumping mentioned here amongst the other factors?

**We added foliage clumping into the list of factors influencing light gradient here:**
**"Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor. The profile shape of this light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types (Fig. 2a-d, Supporting Information Figure S1, Koike et al., 2001)."**

@@

line148: 'Variability in the light environment decreases with height' – I don’t see this except perhaps at z = 0 and even so seems tenuous. Explain.

Fig. 1: In the diagram, stomatal density, leaf angle are separate categories yet VAZ and VOC are not. As far as I know though, there is no biochemical link between VAZ and VOC, so these should be separate shouldn't they?

**We agree to the suggestions. Therefore, to make it more clearer, we added photoprotection as one of the factors in Fig.1 instead of VAZ and VOC in the same line.**

@@

Fig. 2 is a mix of computed and measured quantities. LAD and proportion of sun leaves are both computed quantities, and the basis of these should be clearer. Or at least state 'modelled LAD' and 'calculated proportion of sun leaves'.

**Done**

line165-172: refixation of respired CO2 in the understory isn’t mentioned here, but there have been a number of studies, particularly involving 13C, that have looked at this. Please mention along with at least 1 reference.

**Thank you for this suggestion, we added the following information:**
*"Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions. Particularly, understory seedlings above ground can have highest rates of CO~2~ refixation [Brooks et al. 1997]."*

@@

line206-207 'Leaves are typically warmer than air … '. That is not what is predicted in theory, proven by parts of the simulations in Fig. 3. Why not use 'Leaves can be warmer …'? That would be more consistent.

**Done**

Figure 3: Variables in the inset table need to be stated. I can guess what they are, but this should be explicit.

**Variables in the inset table are now described in the figure caption as: "Biohphysical constants in the the table include: shortwave radiation (swr), wind speed (ws), relative humidity (rh), leaf width (lw), stomatal conductance (gs), air temperature (tair)."**

@@

Figure 3: Humid and drought is not a straight comparison. Well-watered vs. drought would be clearer as there are some assumptions about gs in these conditions that underlie the modelling being done here.

*This is either a semantics issue, or a case where the reviewer is expecting greater realism that what we're actually doing. You should make that clear in the caption (also addresses comment above), and perhaps use "humid" and "dry" (rather than drought).*

*Also note that well-watered would refer to soil moisture, and require modeling water transport through roots and stem. Drought could refer to either soil dryness or atmospheric dryness. In my mind, it's always been the latter.*

**changed to "drought" to "dry" in the figure to refer to atmospheric dryness.**

### 2022-01-18 vertical-thermal-review.Rmd
Adding Tom's micromet section
still working on citations here

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001]. Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers. Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
The implications of this light gradient are that leaf traits vary along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).

>>>>

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor.
The profile shape of light is modified by leaf area density, leaf clumping, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001]. Canopy foliage absorb a large portion of PAR (400-700nm), and selectively filter incident light along the gradient, therefore altering the spectral characteristics of light received in the lower canopy layers. Along with decreasing PAR, there is a decrease in the red (~685-690 nm) to far red (~730-740 nm) ratio of light as it reaches the forest floor, where understories are enriched in near infrared radiation and absorb light in wavelengths of 700-1000 nm [@decastroLightSpectralComposition2000; @poorterLeafOpticalProperties2000].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997].

Steep vertical profiles of wind speed, water vapor, CO~2~ and temperature also occur within canopies, but are more difficult to predict and to generalize on the basis of theoretical first principles than are radiation profiles. Vertical transport at the macroscopic scale of a plant canopy is dominated by turbulent transport rather than by molecular diffusion. Above the canopy, turbulent transport typically mimics diffusion – that is, vertical fluxes of heat, mass and momentum are proportional to their respective vertical gradients, and to transfer coefficients (eddy diffusivities, often symbolized K) that depend on both friction and buoyancy [@penmanWeatherWheatEssay1960]. In this “K-theory” regime, wind speed declines logarithmically with proximity to the ground or canopy surface [@monteithPrinciplesEnvironmentalPhysics2013]. Interaction with a plant canopy also attenuates wind, causing wind speed to decline through the canopy; for example, small plant elements can rapidly dissipate momentum within dense canopies [e.g., @raupachAveragingProceduresFlow1982; @baldocchiSpectralLagcorrelationAnalysis1988]. However, K-theory often fails within canopies, making canopy wind profiles difficult to generalize from first principles [e.g., @denmeadScalarTransportPlant1987; **Raupach 1986**; @meyersModellingPlantCanopy1987; @katulModelingCO2Sources1999; @harmanSimpleUnifiedTheory2007]. For example, in the lower canopy, wind speed may decrease with height despite net downward momentum flux [@shawSecondaryWindSpeed1977], and temperature may increase with height despite upward sensible heat flux (e.g., Raupach 1987). Such counter-gradient transport may arise from the intermittent generation, at the canopy surface, of large wakes or coherent eddy structures that periodically dip down through the canopy, gathering packets of warmed, humidified and CO~2~-enriched or –depleted air from beneath and within the canopy and flushing them to the atmosphere above (Finnigan 1979, Baldocchi and Meyers 1991). These "sweep-eject" events can couple understory conditions more directly to conditions above the canopy than beneath.

Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
Likewise, the mid-canopy experiences a highly variable light environment due to the sun's passage across a dynamically structured canopy surface [@waySunflecksTreesForests2012].
The implications of this light gradient are that leaf traits vary along the vertical gradient [Table 1], and that upper canopy leaves have higher potential photosynthetic rate (see section 4.2), but also greater propensity towards elevated $T_{leaf}$ (see sections 2.2, 4.2).
